---
title: "Problem_Set_2"
author: "Group 2"
date: "2026-02-26"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

\textbf{1.)  Use the bisection method with a hand calculator or computer to find the real root of $x^3 - x - 3 = 0$. Use an error tolerance of $\epsilon = 0.0001$. Graph teh function $f(x) = x^3 - x - 3$ and label the root.}

Solution:

Finding the root using bisection, given $[a,b]=[1,2]$.

```{r, echo=FALSE}
library(cmna)
f_x <- function(x){x^3-x-3}

root <- bisection(f_x, 1, 2, tol=0.0001)
root
```

$\boxed{\text{The root is } 1.671722}$

```{r, echo=FALSE}
curve(f_x, from=0, to=4, col='red')
abline(h=0)

# a and b
abline(v=1, col='blue')
abline(v=2, col='blue')

# root
points(root, 0)
text(root, 2, paste('f(',root,')'))
```

\textbf{2.) The function $f(x) = -3x^3 + 2e^{\frac{x^2}{2}} - 1$ has values of zero near $x=-0.5$ and $x=0.5$.}

  a. What is the derivative of $f$? 
  
  Answer: $\boxed{x(2e^{\frac{x^2}{2}} - 9x)}$
  
  b. If you begin Newton's method at $x = 0$, which root is reached? How many iterations to achieve an error less than $10^{-5}$? 
  
  Answer: If the $x$ is equals to $0$, a root is not reached since in solving for the next $x$ using the equation: $x_{n+1}=x_n-\frac{f(x_n)}{f'(x_n)}$ requires a denominator with nonzero value i.e. $f'(x) \neq 0$. As for the number of iterations, it's not applicable since it simply stops at the first iteration. The plot below plots $f'(0)$.
  
```{r, echo=FALSE}
f_x <- function(x) -3*x^3 + 2*exp(x^2/2) - 1
f_prime <- function(x) 2*x*exp(x^2/2) - 9*x^2

curve(f_x, main="Plotting f(x)", from=-1, to=3, col='red')
abline(h=0)
points(0, f_prime(0))
```
  
  c. Begin Newton's method at another starting point to get the other zero.
  
Answer: Using the starting point $x=1$

```{r, echo=FALSE}
MyNewton <- function(func, d_func, x, eps){
  print(x)
  x_next = x - (func(x) / d_func(x))
  if (abs(x_next - x) <= eps){
    return(x_next)
  }
  
  return(MyNewton(func, d_func, x_next, eps))
}

MyNewton(f_x, f_prime, 1, 0.00001)
```
$\boxed{\text{Therefore, the root is } 0.8568006 \text{ when } x=1}$

\textbf{3.) Use the function from no. 2 and find the root using the secant method where $x_0=0$ and $x_1=1$. Use an error tolerance of $\epsilon = 0.001$.}

```{r, echo=FALSE}
MySecant <- function(f, x_prev, x, tol){
  num <- x - x_prev
  den <- f(x) - f(x_prev)
  
  next_x <- x - f(x) * (num / den)
  
  if(abs(next_x - x) <= tol){
    return(next_x)
  }
  
  print(next_x)
  return(MySecant(f, x, next_x, tol))
}

MySecant(f_x, 0, 1, 0.001)
```

$\boxed{\text{Therefore, the root is } 0.8568007 \text{ when } x_0=0, x_1=1, \text{ and } \epsilon = 0.001 \text{ using the Secant Method}}$

\textbf{4.) Consider the system of equations:}
$$
\begin{cases}
2x + 3y - z = 5 \\
4x - y + 2z = 6 \\
-3x + 2y + z = -4
\end{cases}
$$
 
  a. Present the augmented matrix of the system. Answer:
  
$$
\boxed{\left(
\begin{array}{ccc|c}
2 & 3 & -1 & 5 \\
4 & -1 & 2 & 6 \\
-3 & 2 & 1 & -4
\end{array}
\right)}
$$

  
  b. Solve the system using $Ax=LUx=Ly=b$ and round the final answer to 4 decimal digits.
  
```{r, echo=FALSE}
LU <- function(mat){
  upper <- mat
  n <- nrow(mat)
  lower <- diag(n)
  
  for(col in 1:(n-1)){
    for(row in (col+1):n){
      m <- upper[row, col] / upper[col, col]
      lower[row, col] <- m
      upper[row,] = upper[row,] - m * upper[col,]
      print(sprintf("Row Operation: R_%d ← R_%d - (%.2f)R_%d",
              row, row, m, col))
      print("Upper:")
      print(upper)
      print("Lower:")
      print(lower)
    }
  }
  
  return(list(L = lower, U = upper))
}

A = matrix(c(2, 3, -1, 4, -1, 2, -3, 2, 1), nrow=3, byrow=TRUE)
b = matrix(c(5, 6, -4), nrow=3, byrow=TRUE)
lu_results = LU(A)

L = lu_results$L
U = lu_results$U

# Solving for y using Ly = b
n <- nrow(L)
y <- matrix(0, nrow=3, ncol=1)

for(i in 1:n){
  sum_l <- 0
  if(i > 1){
    for(j in 1:(i-1)){
      sum_l <- sum_l + L[i,j] * y[j,1]
    }
  }
  y[i,1] <- b[i,1] - sum_l
}

# Solving for x using Ux = y
x = matrix(0, nrow=3, ncol=1)

for(i in 3:1){
  sum_u <- 0
  if (i < n) {
    for(j in (i+1):n){
      sum_u <- sum_u + U[i,j] * x[j,1]
    }
  }
  x[i,1] <- (y[i,1] - sum_u) / U[i,i]
}

print(round(y, 4))
print(round(x, 4))
```
$$
\boxed{
\text{Therefore, the final answer is } x=1.6667,y=0.5333,z=-0.0667
}
$$
  c. Find the residual vector if the correct solution is: $x=1,y=2,z=-1$ Answer:
  
  The residual vector can be solved by finding the difference between the correct solution and the answers from the previous question:

$$
\boxed{
\begin{aligned}
x_{residual} &= 1 - 1.6667 = -0.6667 \\
y_{residual} &= 2 - 0.5333 = 1.4667 \\
z_{residual} &= -1 - (-0.0667) = -0.9333
\end{aligned}
}
$$
  
\textbf{5.) Compute the Frobenius norm, maximum column sum, and maximum row sum of the matrix:}
$$
\left(
\begin{array}{ccc}
10.2 & 2.4 & 4.5 \\
-2.3 & 7.7 & 11.1 \\
-5.5 & -3.2 & 0.9
\end{array}
\right)
$$
Answer: 

\textbf{6.) Solve the system of equations given in no.4, starting with the initial vector of $[0,0,0]$:}

  a. Solve using the Jacobi method with 2-digit precision.
  
  b. Solve using Gauss-Seidel method with 2-digit precision.
  
  c. Solve for $\overline{e}$ if the true solution is: $x=1, y=2, z=-1$

