---
title: "Problem Set 2"
author: "Group 2"
date: "March 21, 2026"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

<h4> 1.)  Use the bisection method with a hand calculator or computer to find the real root of $x^3 - x - 3 = 0$. Use an error tolerance of $\epsilon = 0.0001$. Graph the function $f(x) = x^3 - x - 3$ and label the root.} </h4>

<b> Solution: </b>

Finding the root using bisection, given $[a,b]=[1,2]$.

```{r, echo=TRUE}
f_x <- function(x){x^3-x-3}
a <- 1
b <- 2
tol <- 0.0001

bisection_table <- list(
  "Iteration" = c(),
  "a" = c(),
  "b" = c(),
  "c" = c(),
  "f_c" = c(),
  "Err." = c()
)

# Bisection Method Algorithm
iters <- 1
repeat {
  c <- (a + b) / 2
  abs_err <- abs(b - c)
  f_c <- f_x(c)
  
  bisection_table$Iteration <- c(bisection_table$Iteration, iters)
  bisection_table$a <- c(bisection_table$a, a)
  bisection_table$b <- c(bisection_table$b, b)
  bisection_table$c <- c(bisection_table$c, c)
  bisection_table$f_c <- c(bisection_table$f_c, f_c)
  bisection_table$Err. <- c(bisection_table$Err., abs_err)
  
  if(abs_err <= tol){
    break
  }
  
  if(f_x(b) * f_x(c) <= 0){
    a <- c
  } else {
    b <- c
  }
  
  iters <- iters + 1
}

root <- c
bisection_table_df <- as.data.frame(bisection_table)
bisection_table_df
```

$$
\boxed{\text{The root is } 1.6717}
$$

```{r, echo=TRUE}
curve(f_x, from=0, to=4, col='red')
abline(h=0)

# a and b
abline(v=1, col='blue')
abline(v=2, col='blue')

# root
points(root, 0)
text(c, 2, paste('f(',round(root, 4),')'))
```



<h4> 2.) The function $f(x) = -3x^3 + 2e^{\frac{x^2}{2}} - 1$ has values of zero near $x=-0.5$ and $x=0.5$.} </h4>


  a. What is the derivative of $f$? <b> Answer: </b> 
$$
\boxed{
\begin{aligned}
= -3x^3 + 2e^{\frac{x^2}{2}} - 1 \\
= -9x^2 + 2xe^{\frac{x^2}{2}}
\end{aligned}
}
$$
  
  b. If you begin Newton's method at $x = 1$, which root is reached? How many iterations to achieve an error less than $10^{-5}$? <b> Answer: </b>
```{r, echo=TRUE}
f_x <- function(x) { -3*x^3 + 2 * exp(x^2/2) - 1 }
f_prime <- function(x) { 2 * x * exp(x^2 / 2) - 9 * x^2 }
tol <- 0.0001
x_curr <- 1

newton_table <- list(
  "Iteration" = c(),
  "x_curr" = c(),
  "f_x" = c(),
  "f_prime" = c(),
  "x_next" = c(),
  "Err." = c()
)

# Newton's Method Algorithm
iters <- 1
repeat {
  f_curr <- f_x(x_curr)
  f_prime_curr <- f_prime(x_curr)
  
  x_next <- x_curr - f_curr / f_prime_curr
  newton_abs_err <- abs(x_next - x_curr)
  
  newton_table$Iteration <- c(newton_table$Iteration, iters)
  newton_table$x_curr <- c(newton_table$x_curr, x_curr)
  newton_table$f_x <- c(newton_table$f_x, f_curr)
  newton_table$f_prime <- c(newton_table$f_prime, f_prime_curr)
  newton_table$x_next <- c(newton_table$x_next, x_next)
  newton_table$Err. <- c(newton_table$Err., newton_abs_err)
  
  if (newton_abs_err <= tol) {
    break
  }
  
  x_curr <- x_next
  iters <- iters + 1
}

newton_table_df <- as.data.frame(newton_table)
newton_table_df
```

$$
\boxed{\text{Therefore, the root is } 0.8568 \text{ when } x=1}
$$
  
  c. Begin Newton's method at another starting point to get the other zero. <b> Answer: Using the starting point $x=2$ </b>

```{r, echo=TRUE}
x_curr <- 2
newton_table_2 <- list(
  "Iteration" = c(),
  "x_curr" = c(),
  "f_x" = c(),
  "f_prime" = c(),
  "x_next" = c(),
  "Err." = c()
)

iters <- 1
repeat {
  f_curr <- f_x(x_curr)
  f_prime_curr <- f_prime(x_curr)
  
  x_next <- x_curr - f_curr / f_prime_curr
  newton_abs_err <- abs(x_next - x_curr)
  
  newton_table_2$Iteration <- c(newton_table_2$Iteration, iters)
  newton_table_2$x_curr <- c(newton_table_2$x_curr, x_curr)
  newton_table_2$f_x <- c(newton_table_2$f_x, f_curr)
  newton_table_2$f_prime <- c(newton_table_2$f_prime, f_prime_curr)
  newton_table_2$x_next <- c(newton_table_2$x_next, x_next)
  newton_table_2$Err. <- c(newton_table_2$Err., newton_abs_err)
  
  if (newton_abs_err <= tol) {
    break
  }
  
  x_curr <- x_next
  iters <- iters + 1
}

newton_table_2_df <- as.data.frame(newton_table_2)
newton_table_2_df
```

$$
\boxed{\text{Therefore, the root is } 0.8568 \text{ when } x=2}
$$


<h4> 3.) Use the function from no. 2 and find the root using the secant method where $x_0=0$ and $x_1=1$. Use an error tolerance of $\epsilon = 0.001$.} </h4>

```{r, echo=TRUE}
secant_table <- list(
  "x_prev" = c(),
  "x_curr" = c(),
  "x_next" = c(),
  "f_x_curr" = c(),
  "f_x_prev" = c(),
  "Error" = c()
)

MySecant <- function(f, x_prev, x, tol, table) {
  f_x_prev <- f(x_prev)
  f_x <- f(x)
  
  num <- x - x_prev
  den <- f_x - f_x_prev
  
  next_x <- x - f_x * (num / den)
  abs_err <- abs(next_x - x)
  
  table$x_prev <- c(table$x_prev, x_prev)
  table$x_curr <- c(table$x_curr, x)
  table$x_next <- c(table$x_next, next_x)
  table$f_x_prev <- c(table$f_x_prev, f_x_prev)
  table$f_x_curr <- c(table$f_x_curr, f_x)
  table$Error <- c(table$Error, abs_err)
  
  if (abs_err <= tol) {
    return(list(root = next_x, table = table))
  } else {
    return(MySecant(f, x, next_x, tol, table))
  }
}

result <- MySecant(f_x, 0, 1, 0.001, secant_table)

secant_table_df <- as.data.frame(result$table)
secant_table_df
```

$$
\boxed{\text{Therefore, the root is } 0.8568 \text{ when } x_0=0, x_1=1, \text{ and } \epsilon = 0.001 \text{ using the Secant Method}}
$$

<h4> 4.) Consider the system of equations: </h4>
$$
\begin{cases}
2x + 3y - z = 5 \\
4x - y + 2z = 6 \\
-3x + 2y + z = -4
\end{cases}
$$
 
  a. Present the augmented matrix of the system. <b> Answer: </b>
  
$$
\boxed{
\left(
\begin{array}{ccc|c}
2 & 3 & -1 & 5 \\
4 & -1 & 2 & 6 \\
-3 & 2 & 1 & -4
\end{array}
\right)
}
$$

  
  b. Solve the system using $Ax=LUx=Ly=b$ and round the final answer to 4 decimal digits. <b> Answer: </b>
  
```{r, echo=TRUE}
LU <- function(mat){
  upper <- mat
  n <- nrow(mat)
  lower <- diag(n)
  
  for(col in 1:(n-1)){
    for(row in (col+1):n){
      m <- upper[row, col] / upper[col, col]
      lower[row, col] <- m
      upper[row,] = upper[row,] - m * upper[col,]
      print(sprintf("Row Operation: R_%d ← R_%d - (%.2f)R_%d", row, row, m, col))
      print("Upper:")
      print(upper)
      print("Lower:")
      print(lower)
    }
  }
  
  return(list(L = lower, U = upper))
}

A = matrix(c(2, 3, -1, 4, -1, 2, -3, 2, 1), nrow=3, byrow=TRUE)
b = matrix(c(5, 6, -4), nrow=3, byrow=TRUE)
lu_results = LU(A)

L = lu_results$L
U = lu_results$U
n <- nrow(L)

# Solving for y using Ly = b (Forward Substitution)
y <- matrix(0, nrow=3, ncol=1)

for(i in 1:n){
  sum_l <- 0
  if(i > 1){
    for(j in 1:(i-1)){
      sum_l <- sum_l + L[i,j] * y[j,1]
    }
  }
  y[i,1] <- b[i,1] - sum_l
}
print(round(y, 4))

# Solving for x using Ux = y (Backward Substitution)
x = matrix(0, nrow=3, ncol=1)

for(i in 3:1){
  sum_u <- 0
  if (i < n) {
    for(j in (i+1):n){
      sum_u <- sum_u + U[i,j] * x[j,1]
    }
  }
  x[i,1] <- (y[i,1] - sum_u) / U[i,i]
}
print(round(x, 4))
```

$$
\boxed{
\text{Therefore, the final answer is } x=1.6667,y=0.5333,z=-0.0667
}
$$

  c. Find the residual vector if the correct solution is: $x=1, y=2, z=-1$ <b> Answer: </b>

\begin{aligned}

r &= b - Ax \\
r &= 

\left(
\begin{array}{c}
5 \\ 6 \\ -4 
\end{array}
\right)

-

\left(
\begin{array}{ccc}
2 & 3 & -1 \\
4 & -1 & 2 \\
-3 & 2 & 1
\end{array}
\right)

\left(
\begin{array}{c}
1 \\ 2 \\ -1 
\end{array}
\right) \\

r &=

\left(
\begin{array}{c}
5 \\ 6 \\ -4 
\end{array}
\right)

-

\left(
\begin{array}{c}
9 \\ 0 \\ 0 
\end{array}
\right) \\

r &= 

\left(
\begin{array}{c}
-4 \\ 6 \\ -4 
\end{array}
\right)

\end{aligned}

$$
\boxed{\text{Therefore, the residual vector is } (-4, 6, -4)}
$$
  
<h4> 5.) Compute the Frobenius norm, maximum column sum, and maximum row sum of the matrix: </h4>
$$
\left(
\begin{array}{ccc}
10.2 & 2.4 & 4.5 \\
-2.3 & 7.7 & 11.1 \\
-5.5 & -3.2 & 0.9
\end{array}
\right)
$$

<b> Answer: </b>

$$
\begin{aligned}
||A||_f &= \sqrt{(10.2)^2 + (-2.3)^2 + (-5.5)^2 + (2.4)^2 + (7.7)^2 + (-3.2)^2 + (4.5)^2 + (11.1)^2 + (0.9)^2} \\
||A||_f &= \sqrt{104.04 + 5.29 + 30.25 + 5.76 + 59.29 + 10.24 + 20.25 + 123.21 + 0.81} \\
||A||_f &= \sqrt{359.14} \\
||A||_f &= 18.95098942
\end{aligned}
$$

\begin{aligned}
&\text{Maximum Column Sum} \\
\text{Col. 1} &= |10.2| + |-2.3| + |-5.5| = 18.0 \\
\text{Col. 2} &= |2.4| + |7.7| + |-3.2| = 13.3 \\
\text{Col. 3} &= |4.5| + |11.1| + |0.9| = 16.5 \\
\text{MaxColSum} &= max(18.0, 13.3, 16.5) \\
\text{MaxColSum} &= 18.0
\end{aligned}

\begin{aligned}
&\text{Maximum Row Sum} \\
\text{Row. 1} &= |10.2| + |2.4| + |4.5| = 17.1 \\
\text{Row. 2} &= |-2.3| + |7.7| + |11.1| = 21.1 \\
\text{Row. 3} &= |-5.5| + |-3.2| + |0.9| = 9.6 \\
\text{MaxRowSum} &= max(17.1, 21.1, 9.6) \\
\text{MaxRowSum} &= 21.1
\end{aligned}

$$
\boxed{
\begin{aligned}
||A||_f &= 18.95098942 \text{, Frobenius Norm} \\
||A||_\infty &= 18.0 \text{, Maximum column sum} \\
||A||_\infty &= 21.1 \text{, Maximum row sum}
\end{aligned}
}
$$

<h4> 6.) Solve the system of equations given in no.4, starting with the initial vector of $[0,0,0]$: </h4>

  a. Solve using the Jacobi method with 2-digit precision.
```{r, echo=TRUE}
A = matrix(c(2, 3, -1, 4, -1, 2, -3, 2, 1), nrow=3, byrow=TRUE)
b = matrix(c(5, 6, -4), nrow=3, byrow=TRUE)

LUD_jacobi <- function(mat){
  n <- nrow(mat)
  lower <- matrix(0, nrow=n, ncol=n)
  upper <- matrix(0, nrow=n, ncol=n)
  diag  <- matrix(0, nrow=n, ncol=n)
  
  for(i in 1:n){
    diag[i,i] <- mat[i,i]
    j <- 1
    while(j <= (i-1)){
      lower[i,j] <- mat[i,j]
      upper[j,i] <- mat[j,i]
      j <- j + 1 
    }
  }
  
  return(list(L = lower, U = upper, D = diag))
}

LUD_results <- LUD_jacobi(A)

# Lower Triangle
L <- LUD_results$L
L

# Upper Triangle
U <- LUD_results$U
U

# Diagonal
D <- LUD_results$D
D

x <- matrix(0, nrow=3, ncol=1)
tol <- 0.001
for(i in 1:200){
  x <- (-1*solve(D)) %*% (L + U) %*% x + solve(D) %*% b
}
print(round(x, digits = 2))
```

  
  b. Solve using Gauss-Seidel method with 2-digit precision.
  
```{r, echo=TRUE}
x <- matrix(0, nrow=3, ncol=1)
for(i in 1:10){
  #LD <- L + D
  #x <- (-1) %*% solve(LD) %*% U %*% x + solve(LD) %*% b
  #print(x)
}
```
  
  c. Solve for $\overline{e}$ if the true solution is: $x=1, y=2, z=-1$
  

